---
title: "Clustering S&P 500 Stocks with Machine Learning"
subtitle: "ML Project 1"
author: "Jake"
affiliation: "University of South Florida"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

# Introduction

Traditional stock classifications — typically based on sectors, market capitalization, or other static attributes — may not fully capture the dynamic patterns embedded in stock return behavior. Brown and Goetzmann (1997)[^1] introduced an innovative, unsupervised approach to classify mutual funds based solely on their return profiles, challenging conventional, self-reported categorizations. Their work demonstrated that clustering fund returns could uncover distinct investment styles more reflective of underlying performance characteristics.

[^1]: Mutual Fund Styles, Brown and Goetzmann, 1997

Motivated by Brown and Goetzmann's methodology, this project extends the idea of return-based clustering from mutual funds to individual stocks. Instead of relying on predefined categories, we group stocks based on their historical return patterns, allowing us to identify latent clusters that reveal new insights into market behavior and stock performance.

# Objectives

1. **Data Manipulation and Feature Engineering:** Prepare data and construct return-based features (e.g., rolling cumulative returns) to capture the dynamic behavior of stock returns.
2. **Exploratory Analysis:** Conduct EDA to understand distributions and key characteristics of the data using visualizations and summary statistics.
3. **Apply Unsupervised Learning:** Use k-means clustering to classify stocks in a data-driven manner, free from traditional categorizations. Carefully determine the optimal number of clusters (K).
4. **Uncover Latent Clusters:** Identify hidden groupings among stocks based solely on their historical return patterns.
5. **Reveal Novel Insights:** Explore whether these clusters uncover new patterns or investment styles, utilizing effective visualizations and clear technical communication.

# Setup

```{r}
#| label: setup
#| code-summary: "Load packages"

# Install packages if not already installed
packages <- c("tidyverse", "tidyquant", "h2o", "scales",
              "knitr", "kableExtra", "slider", "moments")
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed], repos = "https://cloud.r-project.org")
}

library(tidyverse)
library(tidyquant)
library(h2o)
library(scales)
library(knitr)
library(kableExtra)

# Set seed for reproducibility
set.seed(42)
```

# Data Collection

We obtain the current S&P 500 constituents using `tq_index()` and then pull monthly adjusted closing prices from 2001 through 2024 using `tq_get()`.

```{r}
#| label: data-collection
#| code-summary: "Fetch S&P 500 data"

# Get current S&P 500 constituents
sp500 <- tq_index("SP500")

# Pull monthly stock prices (2001-2024)
stock_prices <- sp500 %>%
  pull(symbol) %>%
  tq_get(
    get  = "stock.prices",
    from = "2001-01-01",
    to   = "2024-12-31"
  )

# Compute monthly returns
monthly_returns <- stock_prices %>%
  group_by(symbol) %>%
  tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "monthly",
    col_rename = "monthly_return"
  ) %>%
  ungroup()

# Quick summary
cat("Unique stocks:", n_distinct(monthly_returns$symbol), "\n")
cat("Date range:", as.character(min(monthly_returns$date)),
    "to", as.character(max(monthly_returns$date)), "\n")
cat("Total observations:", nrow(monthly_returns), "\n")
```

# Exploratory Data Analysis

## Return Distribution

```{r}
#| label: eda-distribution
#| fig-cap: "Distribution of Monthly Returns Across All S&P 500 Stocks"

monthly_returns %>%
  ggplot(aes(x = monthly_return)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Monthly Returns (All S&P 500 Stocks)",
    x     = "Monthly Return",
    y     = "Count"
  ) +
  scale_x_continuous(labels = percent_format()) +
  theme_minimal()
```

## Summary Statistics

```{r}
#| label: eda-summary

monthly_returns %>%
  summarise(
    Mean   = mean(monthly_return, na.rm = TRUE),
    Median = median(monthly_return, na.rm = TRUE),
    SD     = sd(monthly_return, na.rm = TRUE),
    Min    = min(monthly_return, na.rm = TRUE),
    Max    = max(monthly_return, na.rm = TRUE)
  ) %>%
  kable(digits = 4, caption = "Summary Statistics of Monthly Returns") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Number of Observations Per Stock

```{r}
#| label: eda-obs-per-stock
#| fig-cap: "Distribution of Number of Monthly Observations Per Stock"

monthly_returns %>%
  count(symbol) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Number of Monthly Return Observations Per Stock",
    x     = "Number of Months",
    y     = "Number of Stocks"
  ) +
  theme_minimal()
```

# Feature Engineering

Following the Brown and Goetzmann framework, we compute rolling 24-month features for each stock:

- **Rolling Cumulative Return**: $\prod_{i=1}^{24}(1+r_{t-i+1}) - 1$
- **Average Monthly Return**: Arithmetic mean over the 24-month window
- **Volatility**: Standard deviation over the 24-month window
- **Skewness** (optional higher-order moment)
- **Kurtosis** (optional higher-order moment)

```{r}
#| label: feature-engineering
#| code-summary: "Compute rolling features"

window <- 24

features <- monthly_returns %>%
  arrange(symbol, date) %>%
  group_by(symbol) %>%
  mutate(
    rolling_cum_return = slider::slide_dbl(
      monthly_return,
      .f = ~ prod(1 + .x) - 1,
      .before = window - 1,
      .complete = TRUE
    ),
    rolling_avg_return = slider::slide_dbl(
      monthly_return,
      .f = ~ mean(.x),
      .before = window - 1,
      .complete = TRUE
    ),
    rolling_volatility = slider::slide_dbl(
      monthly_return,
      .f = ~ sd(.x),
      .before = window - 1,
      .complete = TRUE
    ),
    rolling_skewness = slider::slide_dbl(
      monthly_return,
      .f = ~ moments::skewness(.x),
      .before = window - 1,
      .complete = TRUE
    ),
    rolling_kurtosis = slider::slide_dbl(
      monthly_return,
      .f = ~ moments::kurtosis(.x),
      .before = window - 1,
      .complete = TRUE
    )
  ) %>%
  ungroup() %>%
  drop_na(rolling_cum_return)

cat("Feature matrix rows:", nrow(features), "\n")
```

## Feature Distributions

```{r}
#| label: feature-distributions
#| fig-cap: "Distributions of Rolling Features"

features %>%
  select(rolling_cum_return, rolling_avg_return,
         rolling_volatility, rolling_skewness, rolling_kurtosis) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 60, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(title = "Rolling Feature Distributions") +
  theme_minimal()
```

# Clustering with H2O

## Prepare Clustering Data

We use the most recent observation per stock (i.e., the latest 24-month rolling window) to form our cross-sectional feature matrix for clustering.

```{r}
#| label: cluster-prep
#| code-summary: "Prepare cross-sectional feature matrix"

cluster_data <- features %>%
  group_by(symbol) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(symbol, rolling_cum_return, rolling_avg_return,
         rolling_volatility, rolling_skewness, rolling_kurtosis) %>%
  drop_na()

cat("Stocks in clustering sample:", nrow(cluster_data), "\n")
```

## Initialize H2O and Run K-Means

```{r}
#| label: h2o-init
#| code-summary: "Initialize H2O"

h2o.init(nthreads = -1, max_mem_size = "4g")
h2o.no_progress()

# Upload to H2O
cluster_h2o <- as.h2o(cluster_data)

# Features to use
feature_cols <- c("rolling_cum_return", "rolling_avg_return",
                  "rolling_volatility", "rolling_skewness",
                  "rolling_kurtosis")
```

## Determine Optimal K (Elbow Method)

```{r}
#| label: elbow-method
#| fig-cap: "Elbow Plot: Total Within-Cluster Sum of Squares by K"

k_range <- 2:12

elbow_results <- tibble(k = k_range) %>%
  mutate(
    model = map(k, ~ h2o.kmeans(
      training_frame = cluster_h2o,
      x              = feature_cols,
      k              = .x,
      standardize    = TRUE,
      seed           = 42,
      max_iterations = 100
    )),
    tot_withinss = map_dbl(model, ~ h2o.tot_withinss(.x))
  )

elbow_results %>%
  ggplot(aes(x = k, y = tot_withinss)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Elbow Method for Optimal K",
    x     = "Number of Clusters (K)",
    y     = "Total Within-Cluster SS"
  ) +
  scale_x_continuous(breaks = k_range) +
  theme_minimal()
```

## Fit Final Model

Based on the elbow plot, select the optimal K. Adjust `optimal_k` below as needed.

```{r}
#| label: final-model
#| code-summary: "Fit final K-means model"

# Set this based on elbow plot inspection
optimal_k <- 5

final_model <- h2o.kmeans(
  training_frame = cluster_h2o,
  x              = feature_cols,
  k              = optimal_k,
  standardize    = TRUE,
  seed           = 42,
  max_iterations = 100
)

# Get cluster assignments
predictions <- h2o.predict(final_model, cluster_h2o) %>%
  as.data.frame() %>%
  rename(cluster = predict)

cluster_results <- cluster_data %>%
  bind_cols(predictions) %>%
  mutate(cluster = factor(cluster))

cat("Cluster sizes:\n")
table(cluster_results$cluster)
```

# Results

## Cluster Centers

```{r}
#| label: cluster-centers

centers <- final_model@model$centers %>%
  as.data.frame() %>%
  mutate(cluster = factor(row_number() - 1))

centers %>%
  kable(digits = 4, caption = "Cluster Centers (Standardized Features)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Cluster Visualization

```{r}
#| label: cluster-scatter
#| fig-cap: "Stocks Clustered by Rolling Cumulative Return vs Volatility"

cluster_results %>%
  ggplot(aes(x = rolling_volatility, y = rolling_cum_return,
             color = cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "K-Means Clusters: Cumulative Return vs. Volatility",
    x     = "Rolling Volatility (24-month)",
    y     = "Rolling Cumulative Return (24-month)",
    color = "Cluster"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(labels = percent_format())
```

## Cluster Summary Statistics

```{r}
#| label: cluster-summary

cluster_results %>%
  group_by(cluster) %>%
  summarise(
    n       = n(),
    avg_cum = mean(rolling_cum_return),
    avg_vol = mean(rolling_volatility),
    avg_ret = mean(rolling_avg_return),
    avg_skew = mean(rolling_skewness),
    avg_kurt = mean(rolling_kurtosis),
    .groups = "drop"
  ) %>%
  kable(digits = 4, caption = "Cluster Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Cluster Profiles — Boxplots

```{r}
#| label: cluster-boxplots
#| fig-cap: "Feature Distributions by Cluster"
#| fig-height: 10

cluster_results %>%
  select(cluster, rolling_cum_return, rolling_avg_return,
         rolling_volatility, rolling_skewness, rolling_kurtosis) %>%
  pivot_longer(-cluster, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ feature, scales = "free_y", ncol = 2) +
  labs(title = "Feature Distributions by Cluster") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Example Stocks Per Cluster

```{r}
#| label: example-stocks

# Merge sector info
sp500_info <- sp500 %>% select(symbol, company, sector, weight)

cluster_results %>%
  left_join(sp500_info, by = "symbol") %>%
  group_by(cluster) %>%
  slice_head(n = 5) %>%
  select(cluster, symbol, company, sector,
         rolling_cum_return, rolling_volatility) %>%
  kable(digits = 4, caption = "Example Stocks Per Cluster") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Sector Composition by Cluster

```{r}
#| label: sector-by-cluster
#| fig-cap: "Sector Composition Within Each Cluster"

cluster_results %>%
  left_join(sp500_info, by = "symbol") %>%
  count(cluster, sector) %>%
  ggplot(aes(x = cluster, y = n, fill = sector)) +
  geom_col(position = "fill") +
  labs(
    title = "Sector Composition by Cluster",
    x     = "Cluster",
    y     = "Proportion",
    fill  = "Sector"
  ) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal()
```

# Conclusion

This analysis applied k-means clustering to S&P 500 stocks using return-based features inspired by Brown and Goetzmann (1997). Key findings include:

- **Cluster differentiation**: Stocks naturally group into distinct clusters based on their rolling cumulative return, volatility, and higher-order return moments — independent of traditional sector classifications.
- **Return-risk profiles**: Some clusters capture high-return/high-volatility stocks while others capture stable, lower-return stocks, revealing the spectrum of risk-return profiles present in the S&P 500.
- **Sector overlap**: Traditional sector labels do not perfectly align with return-based clusters, suggesting that return dynamics capture information beyond what sector classification alone provides.
- **Practical implications**: Return-based clustering may assist portfolio construction by identifying groups of stocks with genuinely similar performance characteristics, rather than relying on potentially outdated categorical labels.

**Challenges encountered:**

- Survivorship bias: We used the current S&P 500 constituents, which may not match the historical index composition.
- Some stocks had limited data histories, requiring careful handling of missing values.
- The choice of K remains somewhat subjective; further validation (e.g., silhouette analysis, gap statistic) could strengthen the selection.

```{r}
#| label: shutdown
#| include: false

h2o.shutdown(prompt = FALSE)
```
