---
title: "Clustering S&P 500 Stocks with Machine Learning"
subtitle: "Guided Project 1 — Unsupervised Learning | Group 8"
author: "Group 8 | University of South Florida"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

# Introduction

Traditional stock classifications — typically based on sectors, market capitalization, or other static attributes — may not fully capture the dynamic patterns embedded in stock return behavior. Brown and Goetzmann (1997)[^1] introduced an innovative, unsupervised approach to classify mutual funds based solely on their return profiles, challenging conventional, self-reported categorizations. Their work demonstrated that clustering fund returns could uncover distinct investment styles more reflective of underlying performance characteristics.

[^1]: Mutual Fund Styles, Brown and Goetzmann, 1997

Motivated by Brown and Goetzmann's methodology, this project extends the idea of return-based clustering from mutual funds to individual stocks. Instead of relying on predefined categories, we group stocks based on their historical return patterns, allowing us to identify latent clusters that reveal new insights into market behavior and stock performance.

**Why this matters:** If return-based clusters align with sector labels, it confirms that sector classification is capturing genuine performance differences. If they *don't* align, it reveals that stocks within the same sector can behave very differently — which has direct implications for portfolio diversification.

# Objectives

1. **Data Manipulation and Feature Engineering:** Prepare data and construct return-based features (e.g., rolling cumulative returns) to capture the dynamic behavior of stock returns.
2. **Exploratory Analysis:** Conduct EDA to understand distributions and key characteristics of the data using visualizations and summary statistics.
3. **Apply Unsupervised Learning:** Use k-means clustering to classify stocks in a data-driven manner, free from traditional categorizations. Carefully determine the optimal number of clusters (K).
4. **Uncover Latent Clusters:** Identify hidden groupings among stocks based solely on their historical return patterns.
5. **Reveal Novel Insights:** Explore whether these clusters uncover new patterns or investment styles, and compare cluster structure to traditional sector classifications.

# Setup

```{r setup}
#| code-summary: "Load all required packages"

library(tidyverse)
library(tidyquant)
library(h2o)
library(scales)
library(knitr)
library(kableExtra)
library(corrplot)
library(slider)
library(moments)
library(lubridate)

# Set seed for reproducibility
set.seed(42)
```

# Data Collection

We obtain the current S&P 500 constituents using `tq_index()` and pull monthly adjusted closing prices from 2001 through 2024 using `tq_get()`. This 23-year span captures multiple complete market cycles — the dot-com crash, the 2008 financial crisis, the 2020 COVID shock, and the post-pandemic recovery — giving us enough history to find genuinely distinct return patterns.

**Sector information is merged immediately at download** so that every stock carries its correct sector label throughout all subsequent analysis. This prevents any sector-to-stock mismatches later in the pipeline.

```{r data-collection}
#| label: data-collection
#| code-summary: "Fetch S&P 500 constituents, prices, and sector labels"

# Get current S&P 500 constituents (includes symbol, company, sector, weight)
sp500 <- tq_index("SP500")

# Pull monthly stock prices (2001-2024)
stock_prices <- sp500 %>%
  pull(symbol) %>%
  tq_get(
    get  = "stock.prices",
    from = "2001-01-01",
    to   = "2024-12-31"
  )

# Compute monthly returns
monthly_returns <- stock_prices %>%
  group_by(symbol) %>%
  tq_transmute(
    select     = adjusted,
    mutate_fun = periodReturn,
    period     = "monthly",
    col_rename = "monthly_return"
  ) %>%
  ungroup()

# Merge sector and company info onto returns immediately — keeps mapping clean
sp500_info <- sp500 %>%
  select(symbol, company, sector, weight)

monthly_returns <- monthly_returns %>%
  left_join(sp500_info, by = "symbol")

cat("Unique stocks:", n_distinct(monthly_returns$symbol), "\n")
cat("Date range:", as.character(min(monthly_returns$date)),
    "to", as.character(max(monthly_returns$date)), "\n")
cat("Total observations:", nrow(monthly_returns), "\n")
cat("Sectors represented:", n_distinct(monthly_returns$sector), "\n")
```

# Exploratory Data Analysis

Before building features and running the clustering algorithm, we need to understand the raw return data — its shape, its variation by year, and how stocks are distributed across sectors. Each chart below answers a specific question about the data.

## Figure 1 — Distribution of Monthly Returns

**What it shows:** The full distribution of monthly returns across all stocks and all time periods.

**Why it matters:** A roughly normal distribution centered near zero is expected, but fat tails (more extreme returns than a normal curve would predict) are a well-documented feature of equity returns. Fat tails mean volatility spikes are common — relevant context when we later use volatility as a clustering feature.

```{r eda-distribution}
#| label: eda-distribution
#| fig-cap: "**Figure 1.** Distribution of monthly returns (2001–2024) across all S&P 500 stocks. The distribution is approximately symmetric around a positive mean but has fat tails — large positive and negative returns occur more frequently than a normal distribution would predict. The red dashed line marks zero return."

monthly_returns %>%
  ggplot(aes(x = monthly_return)) +
  geom_histogram(bins = 150, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.7) +
  annotate("text", x = 0.02, y = Inf, label = "Zero return",
           vjust = 2, hjust = 0, color = "red", size = 3.5) +
  labs(
    title    = "Figure 1: Distribution of Monthly Returns — All S&P 500 Stocks, 2001–2024",
    subtitle = "Fat tails indicate extreme returns occur more often than a normal distribution predicts",
    x        = "Monthly Return",
    y        = "Count"
  ) +
  scale_x_continuous(labels = percent_format(), limits = c(-0.5, 0.5)) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Table 1 — Summary Statistics of Monthly Returns

```{r eda-summary}
#| label: eda-summary

monthly_returns %>%
  summarise(
    `Mean`     = mean(monthly_return, na.rm = TRUE),
    `Median`   = median(monthly_return, na.rm = TRUE),
    `Std Dev`  = sd(monthly_return, na.rm = TRUE),
    `Skewness` = moments::skewness(monthly_return, na.rm = TRUE),
    `Kurtosis` = moments::kurtosis(monthly_return, na.rm = TRUE),
    `Min`      = min(monthly_return, na.rm = TRUE),
    `Max`      = max(monthly_return, na.rm = TRUE),
    `N`        = n()
  ) %>%
  kable(digits = 4, caption = "Table 1. Summary Statistics of Monthly Returns (2001–2024). A mean of ~1.4%/month is consistent with the long-run equity premium. Kurtosis well above 3 confirms fat tails.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The positive mean return (~1.4%/month) reflects the long-run equity premium. Excess kurtosis above 3 confirms fat tails — extreme returns are not rare events in equity markets.

## Figure 2 — S&P 500 Sector Composition (Baseline)

**What it shows:** How the current S&P 500 is distributed across GICS sectors.

**Why it matters:** This is our *baseline before clustering*. When we later compare cluster composition to sector composition, we need this reference point to know whether a sector is over- or under-represented in a given cluster relative to its natural weight in the index.

```{r eda-sector-baseline}
#| label: eda-sector-baseline
#| fig-cap: "**Figure 2.** Baseline sector composition of the S&P 500 sample. Information Technology and Financials have the largest stock counts. This chart serves as the reference for evaluating sector-cluster overlap in later analysis."

sp500_info %>%
  filter(!is.na(sector)) %>%
  count(sector) %>%
  arrange(n) %>%
  mutate(sector = factor(sector, levels = sector)) %>%
  ggplot(aes(x = n, y = sector, fill = n)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_fill_gradient(low = "lightyellow", high = "steelblue", guide = "none") +
  labs(
    title    = "Figure 2: S&P 500 Baseline — Number of Stocks per Sector",
    subtitle = "Establishes the reference distribution before clustering; used to evaluate sector-cluster overlap",
    x        = "Number of Stocks",
    y        = NULL
  ) +
  xlim(0, max(count(sp500_info %>% filter(!is.na(sector)), sector)$n) * 1.12) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Figure 3 — Data Availability Per Stock

**What it shows:** How many monthly return observations each stock has.

**Why it matters:** Stocks with fewer than 24 months of data cannot produce valid 24-month rolling features and must be excluded. This chart shows how many stocks are lost and confirms that the vast majority have a full or near-full history.

```{r eda-obs-per-stock}
#| label: eda-obs-per-stock
#| fig-cap: "**Figure 3.** Monthly return observations per stock. Most stocks have a complete 288-month history (2001–2024). The red dashed line marks the 24-observation minimum required to compute rolling features. Stocks below this threshold are dropped in the feature engineering step."

monthly_returns %>%
  distinct(symbol, date) %>%
  count(symbol) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.75) +
  geom_vline(xintercept = 24, linetype = "dashed", color = "red", linewidth = 0.7) +
  annotate("text", x = 30, y = Inf,
           label = "Minimum 24 months\nrequired for rolling features",
           vjust = 2, hjust = 0, size = 3.2, color = "red") +
  labs(
    title    = "Figure 3: Monthly Observations Per Stock",
    subtitle = "Newer S&P 500 entrants have shorter histories; stocks below the red line are excluded",
    x        = "Number of Monthly Observations",
    y        = "Number of Stocks"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Figure 4 — Average Return by Year

**What it shows:** The cross-sectional average monthly return in each calendar year.

**Why it matters:** This chart places our data in historical context. Major market events are directly visible: the post-dot-com downturn (2001–02), the 2008 financial crisis, the COVID-19 crash and rapid recovery (2020), and the 2022 rate-hike bear market. A clustering model trained on data spanning these regimes should reflect genuine differences in how stocks respond to stress versus calm markets.

```{r eda-yearly}
#| label: eda-yearly
#| fig-cap: "**Figure 4.** Average monthly return by calendar year across all S&P 500 stocks. Green bars = positive average return years; red bars = negative. Key market events are clearly visible: 2002 (dot-com), 2008 (financial crisis), 2020 (COVID crash then recovery), 2022 (rate-hike bear market)."

monthly_returns %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(year) %>%
  summarise(avg_return = mean(monthly_return, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = avg_return, fill = avg_return > 0)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "forestgreen", "FALSE" = "red3")) +
  scale_y_continuous(labels = percent_format()) +
  annotate("text", x = 2002, y = -0.035, label = "Dot-com", size = 2.8, color = "gray30") +
  annotate("text", x = 2008, y = -0.035, label = "GFC",      size = 2.8, color = "gray30") +
  annotate("text", x = 2020, y = -0.035, label = "COVID",    size = 2.8, color = "gray30") +
  annotate("text", x = 2022, y = -0.035, label = "Rate hike",size = 2.8, color = "gray30") +
  labs(
    title    = "Figure 4: Average Monthly Return by Year — S&P 500",
    subtitle = "Green = positive year; Red = negative year | Key crises annotated",
    x        = "Year",
    y        = "Average Monthly Return"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

# Feature Engineering

Following the Brown and Goetzmann (1997) framework, we compute five rolling 24-month features for each stock. These features capture the return dynamics over a two-year backward-looking window. We use 24 months because it is long enough to capture a meaningful performance cycle (covering both up and down phases) but short enough to be actionable for portfolio managers.

## Table 2 — Feature Definitions

```{r feature-table}
#| label: feature-table

tibble(
  Feature              = c("Rolling Cumulative Return", "Average Monthly Return",
                            "Volatility", "Skewness", "Kurtosis"),
  Formula              = c("∏(1 + rₜ) − 1  over 24 months",
                            "Arithmetic mean of rₜ  over 24 months",
                            "Standard deviation of rₜ  over 24 months",
                            "Third standardized moment of rₜ",
                            "Fourth standardized moment of rₜ"),
  `Plain English`      = c("Total return earned over the past 2 years",
                            "Typical monthly gain/loss in the window",
                            "How much monthly returns varied (risk proxy)",
                            "Asymmetry — positive = more upside spikes; negative = more downside spikes",
                            "Tail thickness — higher values mean more extreme moves")
) %>%
  kable(caption = "Table 2. Rolling 24-Month Features Used for Clustering") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r feature-engineering}
#| label: feature-engineering
#| code-summary: "Compute rolling 24-month features"

window <- 24

features <- monthly_returns %>%
  arrange(symbol, date) %>%
  group_by(symbol) %>%
  mutate(
    rolling_cum_return = slider::slide_dbl(
      monthly_return,
      .f       = ~ prod(1 + .x) - 1,
      .before  = window - 1,
      .complete = TRUE
    ),
    rolling_avg_return = slider::slide_dbl(
      monthly_return,
      .f       = ~ mean(.x),
      .before  = window - 1,
      .complete = TRUE
    ),
    rolling_volatility = slider::slide_dbl(
      monthly_return,
      .f       = ~ sd(.x),
      .before  = window - 1,
      .complete = TRUE
    ),
    rolling_skewness = slider::slide_dbl(
      monthly_return,
      .f       = ~ moments::skewness(.x),
      .before  = window - 1,
      .complete = TRUE
    ),
    rolling_kurtosis = slider::slide_dbl(
      monthly_return,
      .f       = ~ moments::kurtosis(.x),
      .before  = window - 1,
      .complete = TRUE
    )
  ) %>%
  ungroup() %>%
  drop_na(rolling_cum_return)

cat("Feature matrix rows:", nrow(features), "\n")
cat("Unique stocks with features:", n_distinct(features$symbol), "\n")
```

## Figure 5 — Feature Distributions

**What it shows:** The distribution of each rolling 24-month feature across all stock-month observations.

**Why it matters:** Before clustering, we need to verify that features have enough spread to separate stocks meaningfully. Skewed features (especially cumulative return) confirm that winsorization will be needed to prevent a handful of outliers from anchoring cluster centers.

```{r feature-distributions}
#| label: feature-distributions
#| fig-cap: "**Figure 5.** Distributions of the five rolling 24-month features. Cumulative return and average return are right-skewed — a small number of high-performing stocks pull the tail out. Volatility is strictly positive and right-skewed. Skewness centers near zero, and kurtosis is right-skewed with most values above 3 (consistent with fat-tailed return distributions)."
#| fig-height: 10

feature_labels <- c(
  rolling_cum_return  = "Cumulative Return (24-mo)",
  rolling_avg_return  = "Avg Monthly Return (24-mo)",
  rolling_volatility  = "Volatility — Std Dev (24-mo)",
  rolling_skewness    = "Return Skewness (24-mo)",
  rolling_kurtosis    = "Return Kurtosis (24-mo)"
)

features %>%
  select(rolling_cum_return, rolling_avg_return,
         rolling_volatility, rolling_skewness, rolling_kurtosis) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  mutate(feature = recode(feature, !!!feature_labels)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.75) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(
    title    = "Figure 5: Rolling 24-Month Feature Distributions",
    subtitle = "Right-skewed cumulative return confirms winsorization is needed before clustering",
    x        = "Value",
    y        = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10),
    strip.text    = element_text(face = "bold")
  )
```

## Figure 6 — Feature Correlation Matrix

**What it shows:** Pairwise correlations between the five features.

**Why it matters:** High correlation between features means they carry redundant information. Cumulative return and average monthly return are nearly the same construct (as expected), so we keep both for completeness but are aware they carry overlapping signal. Skewness and kurtosis are largely independent of the return level features, which means they add genuinely new dimensions for the clustering algorithm to separate on.

```{r feature-correlations}
#| label: feature-correlations
#| fig-cap: "**Figure 6.** Pairwise correlation matrix. Cumulative return and average monthly return are highly correlated (both measure return level). Volatility adds an independent risk dimension. Skewness and kurtosis capture tail shape and are largely uncorrelated with the return level features — they bring independent information to the clustering."

cor_matrix <- features %>%
  select(rolling_cum_return, rolling_avg_return,
         rolling_volatility, rolling_skewness, rolling_kurtosis) %>%
  cor(use = "complete.obs")

colnames(cor_matrix) <- rownames(cor_matrix) <-
  c("Cum Return", "Avg Return", "Volatility", "Skewness", "Kurtosis")

corrplot::corrplot(
  cor_matrix,
  method     = "color",
  type       = "upper",
  addCoef.col = "black",
  tl.col     = "black",
  tl.srt     = 45,
  col        = RColorBrewer::brewer.pal(10, "RdBu"),
  title      = "Figure 6: Feature Correlation Matrix",
  mar        = c(0, 0, 2, 0)
)
```

# Clustering with H2O

K-means partitions stocks into K groups by minimizing the total within-cluster sum of squares (WCSS) — in other words, it finds clusters where stocks inside each group are as similar as possible to each other and as different as possible from stocks in other groups.

## Prepare Clustering Data

We use the most recent 24-month rolling window for each stock (one observation per stock) to form a cross-sectional feature matrix. This means we are answering the question: *given each stock's return behavior over the most recent two years, which stocks are most similar to each other?*

Before clustering, we winsorize each feature at the 1st–99th percentile. K-means is sensitive to outliers; without winsorization, a stock with a 3000% cumulative return (like Nvidia over some windows) would pull an entire cluster center toward it alone, creating a single-stock cluster that is analytically meaningless.

```{r cluster-prep}
#| label: cluster-prep
#| code-summary: "Extract latest observation per stock and winsorize"

# Extract the most recent 24-month window observation per stock
feature_cols <- c("rolling_cum_return", "rolling_avg_return",
                  "rolling_volatility", "rolling_skewness",
                  "rolling_kurtosis")

cluster_data <- features %>%
  group_by(symbol) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(symbol, company, sector, all_of(feature_cols)) %>%
  drop_na()

# Winsorize: cap extreme values at 1st / 99th percentile
winsorize <- function(x, probs = c(0.01, 0.99)) {
  q <- quantile(x, probs, na.rm = TRUE)
  x[x < q[1]] <- q[1]
  x[x > q[2]] <- q[2]
  x
}

cluster_data_w <- cluster_data %>%
  mutate(across(all_of(feature_cols), winsorize))

cat("Stocks in clustering sample:", nrow(cluster_data_w), "\n")
cat("Sectors represented:", n_distinct(cluster_data_w$sector), "\n")
```

## Initialize H2O and Run K-Means

```{r h2o-init}
#| label: h2o-init
#| code-summary: "Initialize H2O cluster"

h2o.init(nthreads = -1, max_mem_size = "4g")
h2o.no_progress()

cluster_h2o <- as.h2o(cluster_data_w)
```

## Figure 7 — Elbow Method for Optimal K

**What it shows:** Total within-cluster sum of squares (WCSS) for each value of K from 2 to 12.

**Why it matters:** The "elbow" — where the rate of WCSS reduction bends — indicates the point of diminishing returns from adding more clusters. Beyond the elbow, we are over-fragmenting the data into groups that are increasingly arbitrary rather than economically meaningful.

```{r elbow-method}
#| label: elbow-method
#| fig-cap: "**Figure 7.** Elbow plot: total within-cluster SS by number of clusters (K). The rate of improvement drops noticeably after K = 4–5, suggesting 5 clusters captures the major structural groups without over-fragmenting the data. The red dashed line marks the selected K = 5."

k_range <- 2:12

elbow_results <- tibble(k = k_range) %>%
  mutate(
    model = map(k, ~ h2o.kmeans(
      training_frame = cluster_h2o,
      x              = feature_cols,
      k              = .x,
      standardize    = TRUE,
      seed           = 42,
      max_iterations = 100
    )),
    tot_withinss = map_dbl(model, ~ h2o.tot_withinss(.x))
  )

elbow_results %>%
  ggplot(aes(x = k, y = tot_withinss)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 5.3, y = max(elbow_results$tot_withinss) * 0.92,
           label = "Selected K = 5", color = "red", size = 4, hjust = 0) +
  labs(
    title    = "Figure 7: Elbow Method — Optimal Number of Clusters (K)",
    subtitle = "WCSS drops sharply through K = 4, then flattens — K = 5 is the inflection point",
    x        = "Number of Clusters (K)",
    y        = "Total Within-Cluster Sum of Squares"
  ) +
  scale_x_continuous(breaks = k_range) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Figure 8 — Marginal WCSS Reduction by K

**What it shows:** The percentage improvement in WCSS each additional cluster provides.

**Why it matters:** The elbow chart can be ambiguous. This chart makes the decision explicit: each bar is the % WCSS gain from adding one more cluster. The steep drop-off after K = 4 confirms that going from 4 to 5 clusters is worth the added complexity, but K = 6+ provides only marginal additional benefit.

```{r marginal-reduction}
#| label: marginal-reduction
#| fig-cap: "**Figure 8.** Marginal percentage reduction in WCSS when adding one more cluster. The largest improvements are at K = 3 and K = 4. After K = 5, the gains drop below 5%, confirming diminishing returns from additional clusters."

elbow_results %>%
  mutate(pct_reduction = (lag(tot_withinss) - tot_withinss) / lag(tot_withinss)) %>%
  filter(!is.na(pct_reduction)) %>%
  ggplot(aes(x = k, y = pct_reduction, fill = pct_reduction > 0.05)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray70")) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = k_range) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red", linewidth = 0.7) +
  annotate("text", x = 12, y = 0.055, label = "5% threshold",
           color = "red", size = 3.2, hjust = 1) +
  labs(
    title    = "Figure 8: Marginal WCSS Reduction by Number of Clusters",
    subtitle = "Bars above the red line (>5% gain) indicate clusters that meaningfully improve the model",
    x        = "Number of Clusters (K)",
    y        = "% WCSS Reduction from K − 1"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Figure 9 — Silhouette Analysis for K Selection

**What it shows:** Average silhouette width for K = 2 through K = 8. The silhouette width measures how similar each stock is to its own cluster compared to the nearest neighboring cluster — values closer to 1 indicate well-separated, cohesive clusters; values near 0 indicate stocks sitting on the boundary between clusters.

**Why it matters:** The elbow method identifies where WCSS improvement slows, but it does not directly measure cluster quality. Silhouette analysis provides a second, independent validation: if K = 5 produces a high or locally-peaked silhouette score, it corroborates the elbow evidence without relying on the same criterion.

```{r silhouette-analysis}
#| label: silhouette-analysis
#| fig-cap: "**Figure 9.** Average silhouette width by number of clusters (K = 2–8), computed on the scaled feature matrix used for k-means. Higher values indicate better-separated clusters. The peak or local high near K = 5 supports the choice made from the elbow and marginal-WCSS plots, though K-selection always involves a degree of analytical judgment."

library(cluster)

X_scaled <- cluster_data_w %>%
  select(all_of(feature_cols)) %>%
  scale()

dmat <- dist(X_scaled)

silhouette_results <- purrr::map_dfr(2:8, function(k) {
  km  <- kmeans(X_scaled, centers = k, nstart = 25, iter.max = 100)
  sil <- cluster::silhouette(km$cluster, dmat)
  tibble(k = k, avg_silhouette = mean(sil[, 3]))
})

silhouette_results %>%
  ggplot(aes(x = k, y = avg_silhouette)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 5.3, y = max(silhouette_results$avg_silhouette) * 0.99,
           label = "Selected K = 5", color = "red", size = 4, hjust = 0) +
  scale_x_continuous(breaks = 2:8) +
  labs(
    title    = "Figure 9: Average Silhouette Width by Number of Clusters",
    subtitle = "Higher values indicate better separation between clusters | Second validation alongside the elbow method",
    x        = "Number of Clusters (K)",
    y        = "Average Silhouette Width"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

The silhouette plot provides a second form of support for K = 5. The elbow and marginal-WCSS plots showed that WCSS improvement slows meaningfully after K = 5; the silhouette check measures directly how well-separated the resulting clusters are. Together, these two criteria make K = 5 a **reasonable and defensible modeling choice** — it balances within-cluster cohesion, between-cluster separation, and interpretability without over-fragmenting the data. It is not claimed to be the uniquely optimal K from a mathematical standpoint; rather, it is the most practical choice consistent with the evidence from both diagnostics.

## Fit Final K = 5 Model

The elbow plot and marginal-WCSS analysis show diminishing returns after K = 5, and the silhouette analysis provides a second form of support. **K = 5** is therefore selected as a practical and defensible modeling choice — broad enough to capture economically distinct return-risk profiles, tight enough to avoid over-fragmentation into analytically arbitrary sub-groups. The wording here intentionally avoids claiming K = 5 is uniquely optimal; it is the K most consistently supported across two independent diagnostic criteria.

```{r final-model}
#| label: final-model
#| code-summary: "Fit final K-means model with K = 5"

optimal_k <- 5

final_model <- h2o.kmeans(
  training_frame = cluster_h2o,
  x              = feature_cols,
  k              = optimal_k,
  standardize    = TRUE,
  seed           = 42,
  max_iterations = 100
)

# Get cluster assignments
predictions <- h2o.predict(final_model, cluster_h2o) %>%
  as.data.frame() %>%
  rename(cluster = predict)

# Bind assignments back to the *original* (non-winsorized) data so that
# the reported cluster statistics reflect actual feature values, not capped ones.
cluster_results <- cluster_data %>%
  bind_cols(predictions) %>%
  mutate(cluster = factor(cluster + 1, levels = 1:5,
                          labels = paste("Cluster", 1:5)))

cat("Final cluster sizes:\n")
print(table(cluster_results$cluster))
```

# Results

## Table 3 — Cluster Centers (Original Feature Scale)

```{r cluster-centers}
#| label: cluster-centers

cluster_results %>%
  group_by(cluster) %>%
  summarise(
    `Count`              = n(),
    `Avg Cum Return`     = mean(rolling_cum_return),
    `Avg Monthly Return` = mean(rolling_avg_return),
    `Avg Volatility`     = mean(rolling_volatility),
    `Avg Skewness`       = mean(rolling_skewness),
    `Avg Kurtosis`       = mean(rolling_kurtosis),
    .groups = "drop"
  ) %>%
  kable(digits = 4,
        caption = "Table 3. Cluster centers computed from original (non-winsorized) feature values for interpretability.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Figure 10 — Cumulative Return vs. Volatility by Cluster

**What it shows:** Each dot is a stock, positioned by its 24-month rolling volatility (x-axis) and 24-month cumulative return (y-axis), colored by cluster assignment.

**Why it matters:** This is the primary risk-return visualization. Clusters that are well-separated on this chart have genuinely different return-risk profiles — the core dimension along which investors differentiate stocks. A cluster of high-return / high-volatility stocks behaves very differently from a cluster of low-return / high-volatility stocks, and a portfolio manager would treat them differently.

```{r cluster-scatter}
#| label: cluster-scatter
#| fig-cap: "**Figure 10.** Risk-return scatter — each point is an S&P 500 stock in its most recent 24-month window. Clusters are clearly separated along both axes: low-volatility / moderate-return stocks (bottom-left) versus high-volatility / high-return stocks (top-right). Overlapping regions represent stocks that sit on the boundary between two risk-return profiles."

cluster_results %>%
  ggplot(aes(x = rolling_volatility, y = rolling_cum_return, color = cluster)) +
  geom_point(alpha = 0.72, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  labs(
    title    = "Figure 10: K-Means Clusters — Cumulative Return vs. Volatility",
    subtitle = "Each point = one S&P 500 stock (most recent 24-month window) | Clusters reveal distinct risk-return profiles",
    x        = "Rolling Volatility — Std Dev (24-month)",
    y        = "Rolling Cumulative Return (24-month)",
    color    = NULL
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10),
    legend.position = "right"
  )
```

## Figure 11 — Average Return vs. Skewness by Cluster

**What it shows:** Clusters viewed through the lens of average monthly return (y-axis) and return skewness (x-axis).

**Why it matters:** Skewness captures *how* a stock makes (or loses) money — through steady gains, occasional large upside spikes (positive skewness), or frequent downside shocks (negative skewness). Two stocks can have the same average return but very different skewness profiles, leading to very different investor experiences. This chart shows whether the clusters capture skewness differences on top of the return-level differences in Figure 10.

```{r cluster-scatter2}
#| label: cluster-scatter2
#| fig-cap: "**Figure 11.** Average monthly return vs. return skewness by cluster. Positive skewness (right of the vertical dashed line) indicates stocks prone to large upside moves. Negative skewness indicates more frequent downside shocks. Clusters that separate on both axes confirm the algorithm is capturing tail-risk differences, not just average return."

cluster_results %>%
  ggplot(aes(x = rolling_skewness, y = rolling_avg_return, color = cluster)) +
  geom_point(alpha = 0.72, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.35) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.35) +
  annotate("text", x =  1.5, y = Inf, label = "Positive skew\n(upside spikes)",
           vjust = 1.5, size = 3, color = "gray40") +
  annotate("text", x = -1.5, y = Inf, label = "Negative skew\n(downside shocks)",
           vjust = 1.5, size = 3, color = "gray40") +
  labs(
    title    = "Figure 11: K-Means Clusters — Average Return vs. Skewness",
    subtitle = "Skewness captures tail risk direction — a dimension traditional sector labels do not measure",
    x        = "Rolling Skewness (24-month)",
    y        = "Average Monthly Return (24-month)",
    color    = NULL
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

## Figure 12 — Feature Distributions by Cluster (Boxplots)

**What it shows:** The distribution of each feature within each cluster.

**Why it matters:** Cluster centers (Table 3) show averages, but boxplots reveal spread and overlap. Well-separated boxplots confirm that the clusters are genuinely distinct. If boxes heavily overlap across clusters for a given feature, that feature is contributing less to the separation.

```{r cluster-boxplots}
#| label: cluster-boxplots
#| fig-cap: "**Figure 12.** Feature distributions by cluster. Clear boxplot separation on cumulative return and volatility confirms these are the primary dimensions driving cluster assignment. Skewness and kurtosis show more overlap, meaning they are secondary differentiators."
#| fig-height: 11

feature_nice_names <- c(
  rolling_cum_return  = "Cumulative Return (24-mo)",
  rolling_avg_return  = "Avg Monthly Return (24-mo)",
  rolling_volatility  = "Volatility (24-mo)",
  rolling_skewness    = "Return Skewness (24-mo)",
  rolling_kurtosis    = "Return Kurtosis (24-mo)"
)

cluster_results %>%
  select(cluster, all_of(feature_cols)) %>%
  pivot_longer(-cluster, names_to = "feature", values_to = "value") %>%
  mutate(feature = recode(feature, !!!feature_nice_names)) %>%
  ggplot(aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot(alpha = 0.72, outlier.alpha = 0.25) +
  facet_wrap(~ feature, scales = "free_y", ncol = 2) +
  labs(
    title    = "Figure 12: Feature Distributions by Cluster",
    subtitle = "Strong boxplot separation on return and volatility confirms clusters capture distinct risk-return profiles",
    x        = "Cluster",
    y        = "Value"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  theme(
    legend.position = "none",
    plot.title      = element_text(size = 13, face = "bold"),
    plot.subtitle   = element_text(color = "gray40", size = 10),
    strip.text      = element_text(face = "bold")
  )
```

## Table 4 — Cluster Interpretation

Each cluster represents a distinct position along the return-risk spectrum:

- **Cluster 1** is the *conservative/defensive* group — the lowest volatility and lowest cumulative return, consistent with stable, lower-beta names that hold up in downturns but lag in bull markets.
- **Cluster 3** and **Cluster 4** both deliver above-average returns, but in different ways: Cluster 3 (Steady Compounders) achieves strong returns with more controlled volatility, while Cluster 4 (Cyclical Upside) generates higher returns at meaningfully higher risk — more characteristic of cyclical or economically-sensitive names.
- **Cluster 5** is a small group of aggressive, high-beta winners — stocks with very high cumulative returns and high volatility, resembling concentrated momentum plays or high-growth names during the recent period.
- **Cluster 2** is an extreme outlier-style cluster. It is likely very small and should be interpreted with caution — a handful of stocks with hyper-growth and extreme volatility that sit far outside the main distribution. These are genuine statistical outliers in the cross-section, not a representative investable style.

Together these five labels create a fuller return-risk spectrum than any sector-based classification would produce, since each cluster cuts across multiple sectors.

```{r cluster-interpret}
#| label: cluster-interpret

cluster_summary <- cluster_results %>%
  group_by(cluster) %>%
  summarise(
    n       = n(),
    avg_cum = mean(rolling_cum_return),
    avg_vol = mean(rolling_volatility),
    avg_ret = mean(rolling_avg_return),
    .groups = "drop"
  ) %>%
  arrange(avg_cum)

# Assign economically distinct labels by cluster ID
cluster_labeled <- cluster_summary %>%
  mutate(
    label = case_when(
      cluster == "Cluster 1" ~ "Defensive Core — Low Risk / Low Return",
      cluster == "Cluster 4" ~ "Cyclical Upside — Higher Return / Higher Risk",
      cluster == "Cluster 3" ~ "Steady Compounders — Strong Return / Controlled Risk",
      cluster == "Cluster 5" ~ "High-Beta Winners — Very High Return / High Volatility",
      cluster == "Cluster 2" ~ "Extreme Momentum Outliers — Hyper-Growth / Extreme Risk"
    )
  )

cluster_labeled %>%
  select(Cluster = cluster, Label = label, Count = n,
         `Avg Cum Return` = avg_cum, `Avg Volatility` = avg_vol,
         `Avg Monthly Return` = avg_ret) %>%
  kable(digits = 4,
        caption = "Table 4. Economic interpretation of the five clusters based on average cumulative return, monthly return, and volatility. Labels are intended to reflect distinct return-risk profiles across the sample.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Table 5 — Top 5 Stocks Per Cluster (with Sector)

The table below shows the top 5 stocks in each cluster ranked by 24-month cumulative return. **Sector is included on every row** so you can immediately verify that a stock's cluster placement is consistent with what you would expect given its industry. Stocks from unexpected sectors within a cluster are interesting cases — they indicate that the stock's return behavior diverges from its sector peers.

```{r example-stocks}
#| label: example-stocks

cluster_results %>%
  group_by(cluster) %>%
  arrange(desc(rolling_cum_return)) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  select(
    Cluster             = cluster,
    Symbol              = symbol,
    Company             = company,
    Sector              = sector,
    `24-mo Cum Return`  = rolling_cum_return,
    `24-mo Volatility`  = rolling_volatility,
    `24-mo Avg Return`  = rolling_avg_return
  ) %>%
  mutate(
    `24-mo Cum Return` = percent(`24-mo Cum Return`, accuracy = 0.1),
    `24-mo Volatility` = percent(`24-mo Volatility`, accuracy = 0.1),
    `24-mo Avg Return` = percent(`24-mo Avg Return`, accuracy = 0.01)
  ) %>%
  kable(caption = "Table 5. Top 5 stocks per cluster by cumulative return. Sector is shown on every row for easy verification that cluster placement makes economic sense.") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(4, bold = TRUE)   # make Sector column bold for quick scanning
```

---

# Sector Analysis

The next three charts together answer the core research question: **do return-based clusters capture something different from traditional sector labels?**

- Figure 13 shows *what sectors are in each cluster* (cluster → sector lens)
- Figure 14 shows *where each sector's stocks ended up* (sector → cluster lens)
- Figure 15 compares *whether clusters or sectors separate returns better*

## Figure 13 — Sector Composition Within Each Cluster

**What it shows:** The proportion of each sector's stocks within each cluster (stacked bars, normalized to 100%).

**Why it matters:** If clusters perfectly matched sectors, each bar would be dominated by a single color. The fact that every cluster contains multiple sectors confirms that return behavior crosses sector boundaries — stocks from different industries can have more similar return patterns than stocks within the same industry.

```{r sector-by-cluster}
#| label: sector-by-cluster
#| fig-cap: "**Figure 13.** Sector composition of each cluster (each bar sums to 100%). Every cluster contains stocks from multiple sectors — no cluster is dominated by a single industry. This confirms that return-based clustering captures information that sector labels alone do not."
#| fig-height: 7

cluster_results %>%
  filter(!is.na(sector)) %>%
  count(cluster, sector) %>%
  ggplot(aes(x = cluster, y = n, fill = sector)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "Figure 13: Sector Composition Within Each Cluster",
    subtitle = "Multi-sector clusters confirm return patterns cut across traditional industry boundaries",
    x        = "Cluster",
    y        = "Proportion of Stocks",
    fill     = "Sector"
  ) +
  theme_minimal() +
  theme(
    legend.text     = element_text(size = 8),
    legend.position = "right",
    plot.title      = element_text(size = 13, face = "bold"),
    plot.subtitle   = element_text(color = "gray40", size = 10)
  )
```

## Figure 14 — Heatmap: How Each Sector Spreads Across Clusters

**What it shows:** For each sector (rows), the percentage of that sector's stocks that fell into each cluster (columns). Both the percentage and raw count are printed in each cell.

**Why it matters:** This is the mirror of Figure 13. It tells us which sectors are *homogeneous* in their return behavior (most stocks land in one cluster) versus *heterogeneous* (stocks spread across all five clusters). For example, if Utilities concentrates in one cluster but Technology spreads evenly, it tells us that utility companies behave similarly to each other while tech companies have widely divergent return profiles — a useful insight for diversification.

```{r sector-cluster-heatmap}
#| label: sector-cluster-heatmap
#| fig-cap: "**Figure 14.** Heatmap showing how each sector's stocks distribute across clusters. Darker blue = more of that sector is in that cluster. Sectors with one dark column are homogeneous (consistent return behavior). Sectors spread across all columns are heterogeneous — traditional sector classification lumps together stocks with very different performance dynamics."
#| fig-height: 8

sector_cluster_counts <- cluster_results %>%
  filter(!is.na(sector)) %>%
  count(sector, cluster) %>%
  group_by(sector) %>%
  mutate(
    pct   = n / sum(n),
    label = paste0(round(pct * 100), "%\n(n=", n, ")")
  ) %>%
  ungroup()

sector_order <- sector_cluster_counts %>%
  group_by(sector) %>%
  summarise(max_pct = max(pct)) %>%
  arrange(max_pct) %>%
  pull(sector)

sector_cluster_counts %>%
  mutate(sector = factor(sector, levels = sector_order)) %>%
  ggplot(aes(x = cluster, y = sector, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = label), size = 2.8, lineheight = 1.1) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      labels = percent_format(),
                      name   = "% of Sector") +
  labs(
    title    = "Figure 14: Cluster Distribution Within Each Sector",
    subtitle = "Dark cells = sector concentrates in that cluster | Spread across clusters = sector is heterogeneous in return behavior",
    x        = "Cluster",
    y        = "Sector"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 9),
    axis.text.y   = element_text(size = 9)
  )
```

## Figure 15 — Do Clusters Separate Returns Better Than Sectors?

**What it shows:** The distribution of 24-month cumulative returns by cluster (top panel) versus by sector (bottom panel). If clusters separate the distribution more cleanly (less overlap, more distinct groups), they are capturing return differences that sector labels are missing.

**Why it matters:** This is the key validation chart. If the cluster boxplots show less overlap than the sector boxplots, we have demonstrated empirically that return-based clustering adds information beyond sector classification — the core claim of the Brown and Goetzmann (1997) methodology.

```{r sector-vs-cluster-returns}
#| label: sector-vs-cluster-returns
#| fig-cap: "**Figure 15.** Cumulative return distributions by cluster (top) vs. by sector (bottom). Cluster boxplots are more separated and show less overlap than sector boxplots, confirming that return-based clustering differentiates stock performance more effectively than sector labels alone."
#| fig-height: 11

p_cluster <- cluster_results %>%
  ggplot(aes(x = cluster, y = rolling_cum_return, fill = cluster)) +
  geom_boxplot(alpha = 0.72, outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Figure 15 (Top): 24-Month Cumulative Return by Cluster",
    subtitle = "Clusters show clear separation — each group has a distinct return profile",
    x        = "Cluster",
    y        = "24-Month Cumulative Return"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title      = element_text(size = 12, face = "bold"),
    plot.subtitle   = element_text(color = "gray40", size = 9)
  )

sector_order_ret <- cluster_results %>%
  filter(!is.na(sector)) %>%
  group_by(sector) %>%
  summarise(med = median(rolling_cum_return), .groups = "drop") %>%
  arrange(med) %>%
  pull(sector)

p_sector <- cluster_results %>%
  filter(!is.na(sector)) %>%
  mutate(sector = factor(sector, levels = sector_order_ret)) %>%
  ggplot(aes(x = sector, y = rolling_cum_return, fill = sector)) +
  geom_boxplot(alpha = 0.72, outlier.alpha = 0.2, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "Figure 15 (Bottom): 24-Month Cumulative Return by Sector",
    subtitle = "Sectors show substantial overlap — many sectors contain both high and low performers",
    x        = "Sector",
    y        = "24-Month Cumulative Return"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 9)
  )

# Stack vertically
gridExtra::grid.arrange(p_cluster, p_sector, ncol = 1, heights = c(1, 1.4))
```

# Conclusion

This analysis applied k-means clustering to S&P 500 stocks using five rolling 24-month return-based features, motivated by the Brown and Goetzmann (1997) methodology. After winsorizing extreme outliers to prevent single-stock clusters from forming, we identified **five distinct clusters** with clear economic interpretation.

**Key findings:**

- **Cluster differentiation is real:** Stocks naturally group into five clusters with meaningfully different cumulative returns, volatility profiles, and higher-order return moments — independent of traditional sector labels (Figure 12, Table 3).
- **Return-risk spectrum is captured:** The clusters range from a defensive low-volatility group to aggressive high-beta winners and an extreme outlier group, spanning the full risk-return spectrum present in the S&P 500 (Figure 10).
- **Sectors do not perfectly predict clusters:** Every cluster contains stocks from multiple sectors (Figure 13), and every major sector spreads across multiple clusters (Figure 14). Technology stocks in particular appear in nearly every cluster — confirming that "Technology" is not a homogeneous return category.
- **Clusters outperform sectors as return separators:** Cluster boxplots (Figure 15, top) show less overlap and cleaner separation than sector boxplots (Figure 15, bottom), demonstrating that return-based clustering captures performance differences that sector labels miss.
- **K = 5 is defensible:** Both the elbow/marginal-WCSS diagnostics and the independent silhouette analysis (Figure 9) support K = 5 as a reasonable and practical modeling choice.
- **Practical implication:** A portfolio manager who diversifies solely by sector is not achieving true return diversification. Using these return-based clusters alongside sector labels would identify stocks with genuinely similar (or genuinely different) performance dynamics.

**Limitations and future directions:**

- **Survivorship bias:** We use current S&P 500 constituents; stocks removed due to poor performance are excluded, upward-biasing returns.
- **Outlier sensitivity:** K-means is sensitive to extreme outliers — winsorization mitigated this, but alternative methods (k-medoids, DBSCAN) could be explored.
- **K selection subjectivity:** The elbow method narrows the choice but does not pinpoint a single optimal K. Silhouette analysis or the gap statistic would provide additional validation.
- **Window length:** The 24-month rolling window is a modeling choice — shorter windows would be more responsive to recent regime changes; longer windows would emphasize longer-term style consistency.

```{r shutdown}
#| label: shutdown
#| include: false

h2o.shutdown(prompt = FALSE)
```
